---
title: "Reproducible Research: Peer Assessment 1"
author: Diana Duncan
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load required libraries and set options

```{r echo=TRUE}
library(ggplot2)
library(lattice)

options(scipen=10)
```

Unzip the activity.zip file and read it into R
```{r echo=TRUE}
if (!file.exists("activity.csv") && file.exists("activity.zip")) 
{
    dataFile <- unzip("activity.zip")
}
activityData <- read.csv("activity.csv", header=TRUE)
```
Transform the date column into a date class

```{r echo=TRUE}
activityData$dateClass <- strptime(activityData$date, format="%Y-%m-%d")
```

## What is mean total number of steps taken per day?

1. Calculate the number of steps taken per day

```{r echo=TRUE}
stepsByDay <- aggregate(steps ~ date, data=activityData, sum)
str(stepsByDay)
```

2. Make a histogram of the total number of steps taken each day

```{r echo=TRUE, fig.width=7, fig.height=7}
histStepsPerDay <- histogram(date ~ steps, data=stepsByDay)
plot(histStepsPerDay)
```

3. Calculate and report the mean and median of the total number of steps taken per day

The mean of the total number of steps per day is `r mean(stepsByDay$steps)` and the median is `r median(stepsByDay$steps)`.

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r echo=TRUE, fig.width=7, fig.height=7}
avgStepsOverInterval <- aggregate(steps ~ interval, data=activityData, FUN=mean)
str(avgStepsOverInterval)

plot(avgStepsOverInterval$interval, avgStepsOverInterval$steps, type="l"
    , main="Average number of steps taken per interval, averaged across all days"
    , xlab="5-minute interval", ylab="Average steps per interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

The 5-minute interval with the maximum number of steps (`r max(avgStepsOverInterval$steps)`) is `r avgStepsOverInterval$interval[which.max(avgStepsOverInterval$steps)]`.

## Imputing missing values

1. Calculate an report the total number of missing values in the dataset (i.e. the total number of rows with ``NA``s)

The total number of missing values for steps in the dataset is `r sum(is.na(activityData$steps))`.

2. Strategy for addressing the missing values: fill in the missing values of the dataset using the mean for the corresponding 5-minute interval.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r echo=TRUE}
imputeNA <- function(x, y) replace(x, is.na(x)
                                   , avgStepsOverInterval$steps[avgStepsOverInterval$interval==y])

activityData$imputedSteps <- mapply(FUN=imputeNA, activityData$steps
                                    , activityData$interval)

activityData2 <- data.frame(steps=as.integer(activityData$imputedSteps)
                            , date=activityData$date, interval=activityData$interval
                            , dateClass=activityData$dateClass)

str(activityData2)
```

4. Make a histogram of the total number of steps taken each day and calculate and report the **mean** and **median** total number of steps taken per day.

```{r echo=TRUE, fig.width=7, fig.height=7}
stepsByDay2 <- aggregate(steps ~ date, data=activityData2, sum)
histStepsPerDay2 <- histogram(date ~ steps, data=stepsByDay2)
plot(histStepsPerDay2)
```

The mean of the total number of steps per day is `r mean(stepsByDay2$steps)` and the median is `r median(stepsByDay2$steps)`.

    + Do these values differ from the estimates from the first part of the assignment?
    
**Yes**
    
    + What is the impact of imputing missing data on the estimates of the total daily number of steps?
    
**It lowers the mean slightly, and the median more significantly**
    
## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.  I elected not to use the ``weekdays()`` function, as it returns the weekday name in the local language.

```{r echo=TRUE}
dayClasses <- c("1"="weekday","2"="weekday","3"="weekday","4"="weekday"
                ,"5"="weekday","6"="weekend","7"="weekend")
activityData2$dayClass <- factor(dayClasses[strftime(activityData2$dateClass
                                                     , format="%u")])
str(activityData2)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r echo=TRUE, fig.width=7, fig.height=7}
avgStepsOverInterval2 <- aggregate(steps ~ interval + dayClass, data=activityData2, FUN=mean)
str(avgStepsOverInterval2)

xyplot(steps ~ interval | dayClass, data=avgStepsOverInterval2, layout= c(1,2)
       , type="l", xlab="Interval", ylab="Number of steps")
```
